---
title: "Assessment 3 - Wrangling"
output: html_document
date: "2025-04-04"
---

```{r}
# install.packages("shinythemes")

library(readr)
library(dplyr)
library(purrr)
library(janitor)
library(lubridate)
library(stringr)
library(ggplot2)
library(skimr)
library(tidyverse)
```



Load all files in the directory
```{r}

all_seasons <- readr::read_csv("Data/all_seasons.csv")
  
shot_files <- list.files("Data/Player_Shots", pattern = "\\.csv$", full.names = TRUE)

print(shot_files)


# Confirm all shot files contain same columns

column_names <- map(shot_files, ~ names(read_csv(.x, n_max = 0))) # read col names
identical_cols <- all(map_lgl(column_names, ~ identical(column_names[[1]], .x))) # check if col names are same

if (!identical_cols) {
  print("Error: Column mismatch detected.")
  column_names
} else {
  print("Success: No column mismatches.")
}

# Combine all shot data files into one main dataset
shots_all <- shot_files %>%
  map(read_csv) %>%
  bind_rows()

```

## Data Cleaning

```{r}
# Standardise column names across datasets
shots_all <- clean_names(shots_all)
all_seasons <- clean_names(all_seasons)

# Check data types
glimpse(shots_all)

# Remove trailing whitespaces before joining
shots_all <- shots_all %>%
  mutate(player_name = str_trim(player_name))

# Convert game_date to date variable
shots_all <- shots_all %>%
  mutate(game_date = mdy(game_date))

# Labels for quarters
shots_all <- shots_all %>%
  mutate(quarter_label = case_when(
    quarter <= 4 ~ paste0("Q", quarter),
    quarter == 5 ~ "OT1",
    quarter == 6 ~ "OT2", 
    quarter == 7 ~ "OT3",
    quarter == 8 ~ "OT4",
    TRUE ~ paste0("OT", quarter - 4)
  ))

# Convert min/sec fields to integers
shots_all <- shots_all %>%
  mutate(
    mins_left = as.integer(mins_left),
    secs_left = as.integer(secs_left)
  )
```

Handle Missing Values
```{r}
colSums(is.na(shots_all))

# Identified position_group data missing for 7930 rows
```


```{r}
position_ref <- read_csv("Data/players.csv") %>%
  clean_names() %>%
  mutate(
    player_name = paste(fname, lname)
  ) %>%
  select(player_name, position_ref = position)



# Join position reference into main dataset for missing values
shots_all <- shots_all %>%
  left_join(position_ref, by = "player_name")

# Fill in missing position groups
shots_all <- shots_all %>%
  mutate(
    position_group = if_else(
      is.na(position_group) & !is.na(position_ref), 
      substr(position_ref, 1, 1), # Take first character eg. G for Guard, F for Forward-Guard, F for Forward
      position_group # keep existing non-NA values
    )
  )

# Clean up and remove position ref
shots_all <- shots_all %>% select(-position_ref)

# Add 5 year bins for help with EDA and visualisations
shots_all <- shots_all %>%
  mutate(
    year_bin = cut(
      season_1,
      breaks = c(2004, 2009, 2014, 2019, 2025), # upper bound 2025 to include 2024
      include.lowest = TRUE, 
      right = FALSE, # left inclusive bins
      labels = c("2004-2008", "2009-2013", "2014-2018", "2019-2024")
    )
  )

shot_counts <- shots_all %>%
  count(year_bin, shot_type)

## Got missing values from 7930 to 4155
```




```{r}
glimpse(all_seasons)

# remove unnecessary index column
all_seasons <- all_seasons %>%
  select(-x1)

all_seasons <- all_seasons %>% 
  mutate(
    draft_year = as.integer(na_if(draft_year, "Undrafted")),
    draft_round = as.integer(na_if(draft_round, "Undrafted")),
    draft_number = as.integer(na_if(draft_number, "Undrafted")), 
    season_start = as.integer(substr(season, 1, 4)) # Extract start year
  )

# Create variable draft status & country group
all_seasons <- all_seasons %>%
  mutate(
    draft_status = if_else(is.na(draft_year), "Undrafted", "Drafted"),
    country_group = if_else(country == "USA", "USA", "International")
  )

# Remove trailing white space before joining

all_seasons <- all_seasons %>%
  mutate(player_name = str_trim(player_name))

colSums(is.na(all_seasons))

# Create new df to show career length for players
career_lengths <- all_seasons %>% 
  group_by(player_name) %>%
  summarise(career_length = n_distinct(season_start))
```


## Merge datasets
```{r}
# Join career lengths to all seasons
all_seasons <- all_seasons %>%
  left_join(career_lengths, by = "player_name")

# Create df for distinct player info
player_info <- all_seasons %>%
  group_by(player_name) %>%
  summarise(
    college = first(na.omit(college)),
    country = first(na.omit(country)), 
    draft_year = first(na.omit(draft_year)),
    career_length = n_distinct(season_start),
    draft_status = first(na.omit(draft_status)),
    country_group = first(na.omit(country_group)),
    .groups = "drop"
  )

# Ensure unqiue values before joining
nrow(player_info) == length(unique(player_info$player_name)) # Return true then good to join

# Join shots_all to all_seasons
nba_data <- shots_all %>%
  left_join(player_info, by = "player_name")


# Export files
write_csv(nba_data, "Data/nba_data.csv")
write_csv(all_seasons, "Data/all_seasons_cleaned.csv")
write_csv(player_info, "Data/player_info.csv")

```



# Exploratory Data Analysis (nba_data - shot_level)
```{r}

#  nba_data EDA
skim(nba_data)
colSums(is.na(nba_data))

```

```{r}
# Shot Type breakdown
ggplot(shot_counts, aes(x = year_bin, y = n, fill = shot_type)) + 
  geom_col(position = "dodge") + 
  labs(title = "Frequency of Shot Types", 
       x = "Season Range",
       y = "Shot Count",
       fill = "Shot Type") +
  theme_minimal()

# Shot Distance Distribution
ggplot(nba_data, aes(x = shot_distance)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") + 
  facet_wrap(~ year_bin) +
  labs(title = "Shot Distance Distribution", 
       x = "Shot Distance (feet)",
       y = "Frequency") +
  theme_minimal()

# USA vs International Shot Style
ggplot(subset(nba_data, !is.na(country_group)), aes(x = country_group, fill = shot_type)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) + 
  labs(title = "Shot Type Distribution: USA vs. International",
       x = "Country Group",
       y = "Proportion",
       fill = "Shot Type")

# Shot Density by Position
ggplot(nba_data, aes(x = loc_x, y = loc_y)) + 
  geom_bin2d(bins = 40) + 
  coord_fixed() + 
  facet_wrap(~position_group) + 
  labs(title = "Shot Density by Position Group")

# Shot Distance by Height
# ggplot(nba_data, aes(x = player_height, y = shot_distance) +
#         geom_point(alpha = 0.2) + 
#         geom_smooth(method = "lm", colour = "red") + 
#         labs(title = "Shot Distance vs Player Height")
       
```

# Exploratory Data Analysis (all seasons)
```{r}
skim(all_seasons)
colSums(is.na(all_seasons))
```
```{r}
# Season stat averages over time
all_seasons %>%
  group_by(season_start) %>%
    summarise(
      avg_pts = mean(pts, na.rm = TRUE),
      avg_reb = mean(reb, na.rm = TRUE),
      avg_ast = mean(ast, na.rm = TRUE),
    ) %>%
  pivot_longer(cols = starts_with("avg"), names_to = "stat", values_to = "value") %>%
  ggplot(aes(x = season_start, y = value, colour = stat)) + 
  geom_line(linewidth = 1) +
  labs(title = "Average PTS, REB, AST Per Season")
  

# Summary table of key variables
print(all_seasons %>% 
  filter(season_start %in% c(2004, 2014, 2022)) %>%
  group_by(season_start) %>%
  summarise(
    avg_height = mean(player_height, na.rm = TRUE),
    avg_weight = mean(player_weight, na.rm = TRUE),
    avg_age = mean(age, na.rm = TRUE),
    avg_ts = mean(ts_pct, na.rm = TRUE),
    avg_usg = mean(usg_pct, na.rm = TRUE),
  ))

# True Shooting Over Time
all_seasons %>%
  group_by(season_start) %>%
  summarise(avg_ts = mean(ts_pct, na.rm = TRUE)) %>%
  ggplot(aes(x = season_start, y = avg_ts)) +
  geom_line(colour = "darkgreen") +
  labs(title = "Average TS% By Season", y = "TS%")

# Career Length Distribution
ggplot(player_info, aes(x = career_length)) +
  geom_histogram(binwidth = 1, fill = "purple") +
  labs(title = "Distribution of Career Lengths")
##### Compare veteran players over time? #####

# 5 to 10 year comparison of USA vs International Players
player_info %>%
  filter(!is.na(country_group)) %>%
  mutate(group = case_when(
    career_length >= 10 ~ "10+ years",
    career_length >= 5 ~ "5-9 years",
    TRUE ~ NA_character_
  ),
  draft_bin = cut(
    draft_year, 
    breaks = seq(1980, 2025, by = 5), 
    labels = paste(seq(1980, 2020, 5), seq(1984, 2024, 5), sep = "-"),
    right = FALSE
  )
  ) %>%
  filter(!is.na(group)) %>%
  count(draft_bin, country_group, group) %>%
  ggplot(aes(x = country_group, y = n, fill = group)) +
  geom_col(position = "dodge") + 
  facet_wrap(~ draft_bin) +
  labs(
    title = "Country Group of Veteran NBA Players Across Eras", 
    x = "Country Group", 
    y = "Count",
    fill = "Career Length"
  ) +
  theme_minimal()
### Add comment somewhere (only 5+ careers shown), also consider filtering NA and before 2019. 


# Height vs. Weight Averages Across Seasons
all_seasons %>%
  filter(!is.na(player_height), !is.na(player_weight)) %>%
  group_by(season_start) %>%
  summarise(
    avg_height = mean(player_height, na.rm = TRUE), 
    avg_weight = mean(player_weight, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(avg_height, avg_weight), 
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = factor(season_start), y = value, fill = metric)) +
  geom_line(size = 1.2) + 
  labs(
    title = "Average Player Height vs. Weight By Season", 
    x = "Season", 
    y = "Value", 
    fill = "Metric"
  ) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))






```



# Create files for dedicated visuals to reduce processing time
```{r}

# Physical trait dfs for archetype visuals
player_phyiscal_traits <- all_seasons %>%
  select(player_name, season_start, player_height, player_weight) %>%
  distinct(player_name, season_start, .keep_all = TRUE) %>%
  filter(!is.na(player_height), !is.na(player_weight))

nba_joined <- nba_data %>%
  left_join(player_phyiscal_traits, by = c("player_name", "season_1" = "season_start"))


# Position group dfs for downstream use
position_group_df <- nba_data %>%
  select(season_1, player_name, position_group) %>%
  distinct()

# Remove 3 duplicates found
position_group_df <- position_group_df %>%
  filter(
    !(player_name == "Chris Johnson" & season_1 == 2013 & position_group == "C"),
    !(player_name == "Marcus Williams" & season_1 == 2008 & position_group == "G"),
    !(player_name == "Marcus Williams" & season_1 == 2009 & position_group == "G")
  )
#duplicate_entries <- position_group_df %>%
#  count(player_name, season_1) %>%
#  filter(n > 1) 
#view(duplicate_entries)


# Visual 1 - Height vs. Shot Distance (animated scatter)
player_archetypes_scatter_data <- nba_joined %>%
  filter(
    !is.na(season_1),
    !is.na(player_height),
    !is.na(shot_distance),
    !is.na(position_group)
  )

player_archetypes_scatter <- player_archetypes_scatter_data %>%
  group_by(season_1, position_group) %>%
  summarise(
    avg_height = mean(player_height, na.rm = TRUE),
    avg_shot_dist = mean(shot_distance, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  arrange(season_1, position_group)

write_csv(player_archetypes_scatter, "Data/Processed/player_archetypes_scatter.csv")


# Visual 2 - Shot Type Composition by Season (2PT vs 3PT)
shot_type_composition <- nba_data %>%
  group_by(season_1, shot_type) %>%
  summarise(shot_count = n(), .groups = "drop")
write_csv(shot_type_composition, "Data/Processed/shot_type_composition.csv")


# Visual 3 - League TS% Over Time 
ts_percentages <- all_seasons %>%
  filter(!is.na(ts_pct), !is.na(country_group)) %>%
  group_by(season_start, country_group) %>%
  summarise(ts_pct = mean(ts_pct, na.rm = TRUE), .groups = "drop")
  
  
league_avg_ts <- all_seasons %>% # league TS% averages
  group_by(season_start) %>%
  summarise(ts_pct = mean(ts_pct), .groups = "drop") %>%
  mutate(country_group = "All")

ts_percentages <- bind_rows(ts_percentages, league_avg_ts) # merge together
ts_percentages <- ts_percentages %>% 
  arrange(season_start, country_group # order data
          )

write_csv(ts_percentages, "Data/Processed/ts_percentages.csv")

# Visual 4 - Usage Across Positions Over Time
usage_positions_df <- all_seasons %>%
  left_join(
    position_group_df, 
    by = c("player_name", "season_start" = "season_1")
  )

usage_positions_df_clean <- usage_positions_df %>%
  select(season_start, player_name, position_group, usg_pct) %>%
  filter(!is.na(position_group), !is.na(usg_pct), !is.na(season_start))
 
usage_percentages <- usage_positions_df_clean %>%
  filter(!is.na(usg_pct), !is.na(position_group)) %>%
  group_by(season_start, position_group) %>%
  summarise(usg_pct = mean(usg_pct, na.rm = TRUE), .groups = "drop")
  
  
league_avg_usg <- usage_positions_df_clean %>% # league usg% averages
  group_by(season_start) %>%
  summarise(usg_pct = mean(usg_pct), .groups = "drop") %>%
  mutate(position_group = "All")

usage_percentages <- bind_rows(usage_percentages, league_avg_usg) # merge together
usage_percentages <- usage_percentages %>% 
  arrange(season_start, position_group)

write_csv(usage_percentages, "Data/Processed/usage_percentages.csv")

# Visual 5 - Physical Profile by Position
height_weight_by_position <- nba_joined %>%
  filter(
    !is.na(season_1),
    !is.na(player_height),
    !is.na(player_weight),
    !is.na(position_group)
  )

player_builds_violin_plot <- height_weight_by_position %>%
  group_by(season_1, position_group) %>%
  summarise(
    avg_height = mean(player_height, na.rm = TRUE),
    avg_weight = mean(player_weight, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  arrange(season_1, position_group)

write_csv(player_builds_violin_plot, "Data/Processed/player_builds_violin_plot.csv")


# Visual 6 - Average Age Demographic of League
avg_age_per_season <- all_seasons %>%
  group_by(season_start) %>%
  summarise(
    avg_age = mean(age, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(season_start, avg_age)

write_csv(avg_age_per_season, "Data/Processed/avg_age_per_season.csv")

# Visual 7 - Career Length 
career_band_player_info <- player_info %>%
  mutate(
    career_band = case_when(
      career_length < 5 ~ "0-4 years",
      career_length < 10 ~ "5-9 years",
      TRUE ~ "10+ years"
    )
  )

career_length_country_draft <- career_band_player_info %>%
  filter(
    !is.na(career_length),
    !is.na(country_group), 
    !is.na(draft_status), 
    !is.na(career_band)
  ) %>%
  group_by(country_group, career_band) %>%
  summarise(
    total_in_band = n(), 
    drafted = sum(draft_status == "Drafted", na.rm = TRUE),
    undrafted = sum(draft_status == "Undrafted", na.rm = TRUE),
    .groups = "drop" 
  ) %>%
  group_by(country_group) %>%
  mutate(
    total_country_players = sum(total_in_band),
    country_pct = total_in_band / total_country_players
  ) %>%
  ungroup() %>%
  select(country_group, career_band, total_in_band, country_pct, drafted, undrafted)

write_csv(career_length_country_draft, "Data/Processed/career_length_country_draft.csv")


# Visual 8 - International Players Over Time
intl_representation <- all_seasons %>%
  group_by(season_start, country_group) %>%
  summarise(player_count = n(), .groups = "drop") %>%
  select(season_start, country_group, player_count)

write_csv(intl_representation, "Data/Processed/intl_representation.csv")
  

# Visual 9 - Player Origin Map
player_origins_season <- all_seasons %>%
  filter(!is.na(season_start), !is.na(country)) %>%
  group_by(season_start, country) %>%
  summarise(player_count = n(), .groups = "drop") %>%
  mutate(view = "Seasonal")

player_origins_season <- player_origins_season %>%
  mutate(season_start = as.character(season_start))

player_origins_overall <- all_seasons %>% 
  filter(!is.na(country)) %>%
  group_by(country) %>%
  summarise(player_count = n(), .groups = "drop") %>%
  mutate(season_start = "All", view = "Overall")

player_origins_map <- bind_rows(player_origins_season, player_origins_overall) %>% # merge together
  select(season_start, country, player_count, view) %>% 
  arrange(season_start, desc(player_count) # order data
          )

# Distinct countries
#player_origins_map_distinct <- player_origins_map %>%
#  distinct(country) %>%
#  arrange(country)

# Clean names to make compatible with natural earth.
player_origins_map <- player_origins_map %>%
  mutate(country = recode(country, 
                          "USA" = "United States of America",
                          "US Virgin Islands" = "United States Virgin Islands",
                          "U.S. Virgin Islands" = "United States Virgin Islands",
                          "Great Britain" = "United Kingdom",
                          "England" = "United Kingdom",
                          "Scotland" = "United Kingdom",
                          "Republic of the Congo" = "Congo",
                          "DRC" = "Democratic Republic of the Congo",
                          "Czech Republic" = "Czechia",
                          "Macedonia" = "North Macedonia",
                          "Russia" = "Russian Federation",
                          "Iran" = "Iran (Islamic Republic of)",
                          "South Korea" = "Korea, Republic of",
                          "Tanzania" = "United Republic of Tanzania",
                          "St. Vincent & Grenadines" = "St. Vincent and the Grenadines",
                          "Saint Lucia" = "St. Lucia", 
                          "Bosnia & Herzegovina" = "Bosnia and Herzegovina",
                          "Bosnia" = "Bosnia and Herzegovina",
                          "Serbia and Montenegro" = "Serbia",
                          "Yugoslavia" = "Serbia",
                          "USSR" = "Russia"
                          ))

write_csv(player_origins_map, "Data/Processed/player_origins_map.csv")

# Visual test - Player Physicals Raw
player_physical_raw <- nba_joined %>%
  filter(!is.na(player_height), !is.na(player_weight), !is.na(season_1), !is.na(position_group)) %>%
  select(season_1, position_group, player_height, player_weight)
write_csv(player_physical_raw, "Data/Processed/player_physical_raw.csv")


# Visual test 2 - physical IQR ranges
height_summary_stats <- player_physical_raw %>%
  group_by(season_1, position_group) %>%
  summarise(
    Q1 = quantile(player_height, 0.25, na.rm = TRUE), 
    Median = median(player_height, na.rm = TRUE), 
    Q3 = quantile(player_height, 0.75, na.rm = TRUE), 
    Min = min(player_height, na.rm = TRUE), 
    Max = max(player_height, na.rm = TRUE),
    .groups = "drop"
  )
write_csv(height_summary_stats, "Data/Processed/height_summary_stats.csv")

weight_summary_stats <- player_physical_raw %>%
  group_by(season_1, position_group) %>%
  summarise(
    Q1 = quantile(player_weight, 0.25, na.rm = TRUE), 
    Median = median(player_weight, na.rm = TRUE), 
    Q3 = quantile(player_weight, 0.75, na.rm = TRUE), 
    Min = min(player_weight, na.rm = TRUE), 
    Max = max(player_weight, na.rm = TRUE),
    .groups = "drop"
  )
write_csv(weight_summary_stats, "Data/Processed/weight_summary_stats.csv")
```
# Additional EDA After file creation
```{r}
# Height vs. Shot Distance
ggplot(nba_joined, aes(x = player_height, y = shot_distance)) + 
  geom_point(alpha = 0.2, colour = "#4AB1F1") + 
  geom_smooth(method = "lm", se = FALSE, colour = "#e76f51") + 
  labs(title = "Shot Distance vs Player Height", 
       x = "Player Height (cm)", 
       y = "Shot Distance (feet)")

# Table of top countries
all_seasons %>%
  group_by(country) %>%
  summarise(player_count = n()) %>%
  arrange(desc(player_count)) %>%
  head(10)


# Distribution of Age by Era
all_seasons %>%
  mutate(year_group = cut(season_start, breaks = c(2000, 2010, 2020, 2025))) %>%
  ggplot(aes(x = age, fill = year_group)) + 
  geom_density(alpha = 0.5) + 
  labs(title = "Distribution of Players by Age")
```



### Explore trends of undrafted players by checking NA vs. values
### Play around with joining inner, and both lefts

# Visual 1 - shot distance trends

nba_data_shot_dist_clean <- nba_data %>%
  filter(!is.na(season_1), !is.na(shot_distance), !is.na(position_group))

shot_distance_trends <- nba_data_shot_dist_clean %>% # position averages
  group_by(season_1, position_group) %>%
  summarise(avg_shot_distance = mean(shot_distance, na.rm = TRUE), .groups = "drop")

league_avg_shots <- nba_data_shot_dist_clean %>% # league averages
  group_by(season_1) %>%
  summarise(avg_shot_distance = mean(shot_distance), .groups = "drop") %>%
  mutate(position_group = "All")

shot_distance_trends <- bind_rows(shot_distance_trends, league_avg_shots) # merge together
shot_distance_trends <- shot_distance_trends %>% 
  arrange(season_1, position_group # order data
          )
write_csv(shot_distance_trends, "Data/Processed/shot_distance_trends.csv")




career_length_country_draft <- career_band_player_info %>%
  filter(
    !is.na(career_length),
    !is.na(country_group), 
    !is.na(draft_status), 
    !is.na(career_band)
  ) %>%
  group_by(country_group, draft_status, career_band) %>%
  summarise(
    avg_career_length = mean(career_length, na.rm = TRUE), 
    player_count = n(),
    .groups = "drop"
  )
